# Conditionals

One aspect of control flow is controlling when and how parts of our code run. This is done by setting actions (or lines of code) to only run when certain conditions are met.

In `R` this is handled by a group of functions under the `if` statements banner.

## `if` statements

The basic `if` statement follows the following syntax:

```{}
if(condition is met){
  
  do this
  
}
```

There are three components to an `if` statement:

*  The `if` special `R` function.
*  The condition to be tested.
*  The code to be evaluated if the condition is met.

In other words, "if this is true do this".

A basic example:

```{r}
x <- 5

if(x > 0){
  
  print("x is a positive number")
  
}

```

```{}
###################
### CHALLENGE 1 ###
###################

Use an if statement to print a suitable message reporting whether there are any people with age of >100 in the titanic dataset. Now do the same for >70. 

HINT: you will need to use the na.rm argument in the any() function.
```

## `if else` statements

So far we've seen code that triggers if the condition is met, but what about if we want to trigger something different if the condition is not met? This is where the `else` statement works in conjunction with the `if` statement:

```{r}
x <- 5

if(x < 0){
  
  print("x is less than zero")
  
} else {
  
  print("x is greater than or equal to zero")
  
}
```

What if you have more than a two category condition? Instead of "if true do this" we might have "if A do this, if B do this, if C do this". 

```{r}
x <- 5

if(x < 0){
  
  print("x is less than zero")
  
} else if(x == 0){
  
  print("x equals zero")
  
} else if(x > 0){
  
  print("x is greater than zero")
  
}
```

```{}
###################
### CHALLENGE 2 ###
###################

Randomly sample one passenger from the titanic dataset. Then construct an if else statement to print a message about whether the passenger is younger than 20, between 20 and 50, or over 50 years old.
```

## The `ifelse()` function

You may have noticed that the `if` statement performs a single test. This means that you can't apply it repeatedly over an `R` object like a vector. The `ifelse()` function is vectorised so it can handle vector inputs. The three inputs to this function are the `test` to be applied to each element of the vector, the code to happen if the test returns TRUE (`yes`), and the code to happen if the test returns FALSE (`no`).

```{r}
x <- 1:10

ifelse(test = x %% 2 == 0, 
       yes = "even", 
       no = "odd")
```

```{}
###################
### CHALLENGE 3 ###
###################

Add a new column to the titanic dataset called "Adult" that is a binary variable indicating whether that passenger is an 18 or older.
```

## Challenge Answers

```{r, eval = FALSE}
###################
### Challenge 1 ###
###################

if(any(titanic$Age > 100,
       na.rm = TRUE)){
  
  print("There are passengers older than 100")
  
}

if(any(titanic$Age > 70,
       na.rm = TRUE)){
  
  print("There are passengers older than 70")
  
}
```

```{r, eval = FALSE}
###################
### CHALLENGE 2 ###
###################

passenger <- titanic[sample(1:nrow(titanic), 1), ]

if(is.na(passenger$Age)){
  
  print("The passenger has no recorded age")
  
} else if(passenger$Age < 20){
  
  print("The passenger is less than 20 years old")

} else if(passenger$Age >= 20 & <= 50){
  
  print("The passenger is between 20 and 50 years old")

} else if(passenger$Age > 50){
  
  print("The passenger is greater than 50 years old")

}
```

```{r, eval = FALSE}
###################
### CHALLENGE 3 ###
###################

titanic$Adult <- ifelse(titanic$Age < 18,
                        yes = 0,
                        no = 1)
```

