# Loops

The second aspect of control flow is being able to repeat the same set of operations for a series of inputs. This is achieved by using loops to iteratively repeat your code until some condition is met.

There are two types of loops in `R`: `for` loops and `while` loops. The most common is `for` loops where you repeat your code *for* each element in a series of inputs. `while` loops will repeat over the same set of code *while* a condition is met.

## `for` loops

The basic syntax of a for loop is this:

```{}
for(iterator in some_things){

  do this
  
}
```

A basic example:

```{r}
for(i in 1:10){
  
  print(i)
  
}
```

```{}
###################
### CHALLENGE 1 ###
###################

Basic for loop challenge
```

### Nested `for` loops

For situations where you need to loop over the different combinations of multiple iterators you can nest `for` loops within each other. For example:

```{r}
for(i in 1:5){
  
  for(j in c("a", "b", "c", "d", "e")){
    
    print(paste(i, j))
  }
}
```

```{}
###################
### CHALLENGE 2 ###
###################

Nested for loop challenge
```

Something about growing output?

## `while` loops

`while` loops have a more niche application, but are really useful for when you don't know how long it will take you to get to the end. A great application example is population viability simulations where you keep progressing through time until the population dies out.

The basic syntax looks life this:

```{}
while(condition is true){

  do this
  
}
```

For example:

```{r}
x <- 10

while(x > 0){
  
  x <- x - 1
  
  print(x)
  
}
```

```{}
###################
### CHALLENGE 3 ###
###################

Use a while loop to simulate a population decline. Start with a population of 100 and then use the sample() function to have it decline between 1-5 individuals per time step.
```
