# Loops

The second aspect of control flow is being able to repeat the same set of operations for a series of inputs. This is achieved by using loops to iteratively repeat your code until some condition is met.

There are two types of loops in `R`: `for` loops and `while` loops. The most common is `for` loops where you repeat your code *for* each element in a series of inputs. `while` loops will repeat over the same set of code *while* a condition is met.

## `for` loops

The basic syntax of a for loop is this:

```{}
for(iterator in some_things){

  do this
  
}
```

A basic example:

```{r}
for(i in 1:10){
  
  print(i)
  
}
```

```{}
###################
### CHALLENGE 4 ###
###################

For each of the passenger classes in the titanic dataset calculate the mean survival rate of that class
```

### Nested `for` loops

For situations where you need to loop over the different combinations of multiple iterators you can nest `for` loops within each other. For example:

```{r}
for(i in 1:5){
  
  for(j in c("a", "b", "c", "d", "e")){
    
    print(paste(i, j))
  }
}
```

```{}
###################
### CHALLENGE 5 ###
###################

For each combination of passenger class and sex in the titanic dataset calculate the mean survival rate
```

### Growing outputs

A strength of loops is that we can use them to perform the same analysis over and over. However there are two ways of storing the output. The first is known as the *balloon method* and involves continually expanding the size of your output. For example:

```{r}
output_vector <- c()

for(i in 1:10){
  
  output_vector <- c(output_vector,
                     sample(1:5, 1))
  
  print(output_vector)
  
}
```

The alternative is the *bottle method* which involves creating an empty output of the correct size and then filling it piece by piece. The *balloon method* is easier to set up but the *bottle method* is more computationally efficient as it does not involve re-writing the output everytime you add to it. For example:

```{r}
output_vector <- rep(0, 10)

for(i in 1:10){
  
  output_vector[i] <- sample(1:5, 1)

  print(output_vector)
  
}
```

```{}
###################
### CHALLENGE 6 ###
###################

Use the system.time() function to compare how long it takes to use the balloon and bottle methods to create a vector 10,000 elements long
```

## `while` loops

`while` loops have a more niche application, but are really useful for when you don't know how long it will take you to get to the end. A great application example is population viability simulations where you keep progressing through time until the population dies out.

The basic syntax looks life this:

```{}
while(condition is true){

  do this
  
}
```

For example:

```{r}
x <- 10

while(x > 0){
  
  x <- x - 1
  
  print(x)
  
}
```

```{}
###################
### CHALLENGE 7 ###
###################

Use a while loop to simulate a population decline. Start with a population of 100 and then use the sample() function to have it decline between 1-5 individuals per time step.
```

## Challenge Answers

```{r, eval = FALSE}
###################
### CHALLENGE 4 ###
###################

for(class in unique(titanic$Pclass)){
  
  mean_surv <- mean(titanic[titanic$Pclass == class, "Survived"],
                    na.rm = TRUE)
  
  print(paste("The mean survival rate of class", class, "was", mean_surv))
  
}
```

```{r, eval = FALSE}
###################
### CHALLENGE 5 ###
###################

for(class in unique(titanic$Pclass)){
  
  for(sex in unique(titanic$Sex)){
    
  mean_surv <- mean(titanic[titanic$Pclass == class &
                              titanic$Sex == sex, 
                            "Survived"],
                    na.rm = TRUE)
  
  print(paste("The mean survival rate of", sex, "passengers in class", class, "was", mean_surv))
    
  }
}
```

```{r, eval = FALSE}
### Balloon method

output_vector <- c()

system.time(
  
  for(i in 1:10000){
    
    output_vector <- c(output_vector,
                       1)
    
  }
)

### Bottle Method

output_vector <- vector(length = 10000)

system.time(
  
  for(i in 1:10000){
    
    output_vector[i] <- 1
    
  }
)
```

```{r, eval = FALSE}
###################
### CHALLENGE 7 ###
###################

population <- 100

while(population > 0){
  
  population <- population - sample(1:5, 1)
  
  print(population)
}
```

